Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-03-05 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex
Fix table overflow
add paragrah about sampling issue due to large m

Furthermore, from Figure~\ref{fig7_lscale_comparison} left it can be seen that using large number of basis functions relative to the wigglyness of the function to be learned (i.e., true length-scale) causes overestimating the underlying functions (see Figure~\ref{fig8_overestimated_posterios} as an illustration of the overestimating effect as a function of $m$ in a moderate smooth dataset) and, consequently, on the length-scale estimate too (Figure \ref{fig7_lscale_comparison} left).

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{fig8_overestimated_posterios.png}
\caption{Overestimating effect on posterior functions when using too many number of basis functions relative to the wigglyness of the function to be learned. The dashed red line represents the true function.}
  \label{fig8_overestimated_posterios}
\end{figure}

@@

\begin{table*}
\centering
\begin{tabular}{ c c c c c c c c c c | c c}
\arrayrulecolor{gray}\hline \\[-3mm]
$\ell_{\text{GP}}$ & iter. & $\ell$ & $c$ & $m$ & $\hat{\ell}$ & $\hat{\ell} + 0.01 \geq \ell$ & RMSE & $R^2$ & ELPD & RMSE* & \\ 
%\arrayrulecolor{lightgray}\hline \\[-1mm]
%\multirow{5}{*}{ 0.05} 
% & 1 & 0.50 & 1.60 & 6 & 0.127 & FALSE & 0.800 & 0.264 & -0.980 & 0.748 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{fig9_diagnostic_1.png}}\\
% & 2 & 0.127 & 1.20 & 17 & 0.061 & FALSE & 0.427 & 0.779 & -0.396 & 0.330 & \\
% & 3 & 0.061 & 1.20 & 35 & 0.059 & TRUE & 0.264 & 0.908 & 0.008 & 0.063 & \\
% & 4 & 0.053 & 1.20 & 40 & 0.059 & TRUE & 0.254 & 0.915 & 0.088 & 0.029 & \\
% \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{5}{*}{ 0.08} 
 & 1 & 0.50 & 1.60 & 6 & 0.17 & FALSE & 0.45 & 0.61 & -0.40 & 0.36 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{fig9_diagnostic_2.png}}\\
 & 2 & 0.17 & 1.20 & 13 & 0.07 & FALSE & 0.36 & 0.74 & -0.18 & 0.24 & \\
 & 3 & 0.07 & 1.20 & 31 & 0.08 & TRUE & 0.24 & 0.87 & 0.20 & 0.01 & \\
 & 4 & 0.06 & 1.20 & 36 & 0.08 & TRUE & 0.24 & 0.87 & 0.21 & 0.01 & \\
 \\[1mm]
 \arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.13} 
 & 1 & 0.50 & 1.60 & 6 & 0.25 & FALSE & 0.46 & 0.63 & -0.49 & 0.33 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{fig9_diagnostic_3.png}}\\
 & 2 & 0.25 & 1.20 & 9 & 0.15 & FALSE & 0.35 & 0.78 & -0.14 & 0.15 & \\
 & 3 & 0.15 & 1.20 & 15 & 0.15 & TRUE & 0.32 & 0.81 & -0.07 & 0.08 & \\
 & 4 & 0.11 & 1.20 & 20 & 0.12 & TRUE & 0.30 & 0.83 & -0.01 & 0.01 & \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.25} 
 & 1 & 0.50 & 1.60 & 6 & 0.24 & FALSE & 0.37 & 0.40 & -0.15 & 0.21 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{fig9_diagnostic_4.png}}\\
 & 2 & 0.24 & 1.20 & 9 & 0.26 & TRUE & 0.30 & 0.60 & -0.09 & 0.02 & \\
 & 3 & 0.25 & 1.20 & 14 & 0.26 & TRUE & 0.29 & 0.62 & -0.05 & 0.01 & \\
 \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.54} 
 & 1 & 0.50 & 1.60 & 6 & 0.38 & FALSE & 0.27 & 0.86 & 0.03 & 0.01 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{fig9_diagnostic_5.png}}\\
 & 2 & 0.38 & 1.20 & 6 & 0.47 & TRUE & 0.28 & 0.86 & 0.03 & 0.05 & \\
 & 3 & 0.24 & 1.50 & 11 & 0.53 & TRUE & 0.27 & 0.86 & 0.04 & 0.01 & \\
 \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.80} 
 & 1 & 1.00 & 3.20 & 6 & 0.52 & FALSE & 0.29 & 0.92 & 0.13 & 0.03 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{fig9_diagnostic_6.png}}\\
 & 2 & 0.52 & 1.65 & 6 & 0.98 & TRUE & 0.29 & 0.92 & 0.11 & 0.03 & \\
 & 3 & 0.50 & 3.12 & 11 & 0.77 & TRUE & 0.28 & 0.92 & 0.11 & 0.01 & \\
 \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{2}{*}{ 1.40} 
 & 1 & 1.00 & 3.20 & 6 & 1.02 & TRUE & 0.28 & 0.56 & 0.14 & 0.01 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{fig9_diagnostic_7.png}}\\
 & 2 & 0.52 & 3.27 & 11 & 1.23 & TRUE & 0.28 & 0.57 & 0.14 & 0.00 & \\
 \\
 \\
 \\[1mm]
 \arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{2}{*}{ 2.94} 
 & 1 & 0.50 & 1.60 & 6 & 1.19 & TRUE & 0.34 & 0.45 & -0.10 & 0.06 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{fig9_diagnostic_8.png}}\\
 & 2 & 0.61 & 3.81 & 11 & 2.58 & TRUE & 0.34 & 0.45 & -0.07 & 0.01 & \\
 \\
 \\
 \\[1mm]
\arrayrulecolor{gray}\hline
\end{tabular}
\caption{diagnostic on various data sets with varying smoothness. RMSE* refers to the root mean squared error computed with the exact GP model as the reference.}
  \label{tab1_diagnostic}
\end{table*}

@@

\begin{table}
\centering
\begin{tabular}{ccccccccc}
\arrayrulecolor{gray}\hline \\[-3mm]
iter. & f & $\ell$ & $c$ & $m$ & $\hat{\ell}$ & $\hat{\ell} + 0.01 \geq \ell$ & RMSE & ELPD \\ 
\arrayrulecolor{lightgray}\hline \\[-2mm]
   & $f_1$ & 0.52 & 1.20 & 7  & 0.53 & FALSE &      &  	   \\
 1 & $f_2$ & 0.50 & -    & 8  & 0.34 & FALSE & 0.32 & 0.05 \\
   & $f_3$ & 0.50 & -    & 8  & 1.62 & TRUE  &      &  	   \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
   & $f_1$ & 0.50 & 1.20 & 8  & 0.54 & TRUE  &      &  	   \\
 2 & $f_2$ & 0.34 & -    & 11 & 0.29 & FALSE & 0.32 & 0.06 \\
   & $f_3$ & 0.29 & -    & 13 & 1.53 & TRUE  &      &  	   \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
   & $f_1$ & 0.28 & 1.20 & 13 & 0.33 & TRUE  &      &  	   \\
 3 & $f_2$ & 0.29 & -    & 13 & 0.24 & FALSE & 0.30 & 0.12 \\
   & $f_3$ & 0.29 & -    & 13 & 1.64 & TRUE  &      &  	   \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
   & $f_1$ & 0.28 & 1.20 & 18 & 0.23 & TRUE  &      &  	   \\
 4 & $f_2$ & 0.24 & -    & 16 & 0.23 & TRUE  & 0.30 & 0.15 \\
   & $f_3$ & 0.29 & -    & 13 & 1.68 & TRUE  &      &  	   \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
   & $f_1$ & 0.16 & 1.20 & 23 & 0.20 & TRUE  &      &  	   \\
 5 & $f_2$ & 0.24 & -    & 16 & 0.23 & TRUE  & 0.29 & 0.18 \\
   & $f_3$ & 0.29 & -    & 13 & 1.63 & TRUE  &      &  	   \\[1mm]
\arrayrulecolor{gray}\hline
\end{tabular}
\caption{Iterative steps in HSGP model fitting of \textit{Birthday data} case study. The diagnosis procedure updating the parameters minimum $\ell$, $c$, $m$, estimated $\hat{\ell}$ and diagnostic $\hat{\ell} + 0.01 \geq \ell$ is applied on the three underlying functions $f_1$, $f_2$ and $f_3$. Notice that $f_2$ and $f_3$ do not have boundary factor $c$ as they use periodic covariance functions. RMSE and ELPD are evaluated for the model underlying function $\mu=f_1 + f_2 + f_3 + f_4$ and sample data. The full diagnosis process is applied until two \textit{trues} are achieved for each function.}
  \label{tab_caseI}
\end{table}

>>>>

If much bigger $m$ than necessary is used (i.e., too much large $m$ relative to the wigglyness of the function to be learned), there can be computational challenges. In Figure~\ref{fig7_lscale_comparison}, we see that increasing $m$ much beyond what is needed, is leading to underestimating the lengthscale. Examining the convergence diagnostics revealed problems in exploring larger length-scales, which is likely due to the posterior having a difficult funnel shape \citep[see, e.g.,][]{Betancourt+Girolami:2019}. A largely underestimation of the length-scale leads to overestimating the underlying functions (see Figure~\ref{fig8_overestimated_posterios} as an illustration of the overestimating effect as a function of $m$ in a moderate smooth dataset). Increasing the boundary factor $c$ (more than that recommended in equations~\eqref{sec_num_equations}) reduces the funnellness of the posterior and the sampling problems in exploring larger length-scales mostly disappear and so underestimating the length-scale.

Following our recommended diagnostic tool is a safety procedure to avoid using too much large $m$ that needed, as it allows us to select optimal values for $m$ and $c$ in each iteration.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth, trim = 0mm 0mm 0mm 0mm, clip]{fig8_overestimated_posterios.png}
\caption{Overestimation of the data points when using too many number of basis functions relative to the wigglyness of the data points. The dashed red line represents the true underlying function.}
  \label{fig8_overestimated_posterios}
\end{figure}

@@

\begin{table*}
\centering
\setlength{\tabcolsep}{5pt}
\begin{tabular}{ c c c c c c c c c c | c c}
\arrayrulecolor{gray}\hline \\[-3mm]
$\ell_{\text{GP}}$ & iter. & $\ell$ & $c$ & $m$ & $\hat{\ell}$ & $\hat{\ell} + 0.01 \geq \ell$ & RMSE & $R^2$ & ELPD & RMSE* & \\ 
%\arrayrulecolor{lightgray}\hline \\[-1mm]
%\multirow{5}{*}{ 0.05} 
% & 1 & 0.50 & 1.60 & 6 & 0.127 & FALSE & 0.800 & 0.264 & -0.980 & 0.748 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{fig9_diagnostic_1.png}}\\
% & 2 & 0.127 & 1.20 & 17 & 0.061 & FALSE & 0.427 & 0.779 & -0.396 & 0.330 & \\
% & 3 & 0.061 & 1.20 & 35 & 0.059 & TRUE & 0.264 & 0.908 & 0.008 & 0.063 & \\
% & 4 & 0.053 & 1.20 & 40 & 0.059 & TRUE & 0.254 & 0.915 & 0.088 & 0.029 & \\
% \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{5}{*}{ 0.08} 
 & 1 & 0.50 & 1.60 & 6 & 0.17 & FALSE & 0.45 & 0.61 & -0.40 & 0.36 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{fig9_diagnostic_2.png}}\\
 & 2 & 0.17 & 1.20 & 13 & 0.07 & FALSE & 0.36 & 0.74 & -0.18 & 0.24 & \\
 & 3 & 0.07 & 1.20 & 31 & 0.08 & TRUE & 0.24 & 0.87 & 0.20 & 0.01 & \\
 & 4 & 0.06 & 1.20 & 36 & 0.08 & TRUE & 0.24 & 0.87 & 0.21 & 0.01 & \\
 \\[1mm]
 \arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.13} 
 & 1 & 0.50 & 1.60 & 6 & 0.25 & FALSE & 0.46 & 0.63 & -0.49 & 0.33 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{fig9_diagnostic_3.png}}\\
 & 2 & 0.25 & 1.20 & 9 & 0.15 & FALSE & 0.35 & 0.78 & -0.14 & 0.15 & \\
 & 3 & 0.15 & 1.20 & 15 & 0.15 & TRUE & 0.32 & 0.81 & -0.07 & 0.08 & \\
 & 4 & 0.11 & 1.20 & 20 & 0.12 & TRUE & 0.30 & 0.83 & -0.01 & 0.01 & \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.25} 
 & 1 & 0.50 & 1.60 & 6 & 0.24 & FALSE & 0.37 & 0.40 & -0.15 & 0.21 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 5mm, clip]{fig9_diagnostic_4.png}}\\
 & 2 & 0.24 & 1.20 & 9 & 0.26 & TRUE & 0.30 & 0.60 & -0.09 & 0.02 & \\
 & 3 & 0.25 & 1.20 & 14 & 0.26 & TRUE & 0.29 & 0.62 & -0.05 & 0.01 & \\
 \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.54} 
 & 1 & 0.50 & 1.60 & 6 & 0.38 & FALSE & 0.27 & 0.86 & 0.03 & 0.01 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{fig9_diagnostic_5.png}}\\
 & 2 & 0.38 & 1.20 & 6 & 0.47 & TRUE & 0.28 & 0.86 & 0.03 & 0.05 & \\
 & 3 & 0.24 & 1.50 & 11 & 0.53 & TRUE & 0.27 & 0.86 & 0.04 & 0.01 & \\
 \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{3}{*}{ 0.80} 
 & 1 & 1.00 & 3.20 & 6 & 0.52 & FALSE & 0.29 & 0.92 & 0.13 & 0.03 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{fig9_diagnostic_6.png}}\\
 & 2 & 0.52 & 1.65 & 6 & 0.98 & TRUE & 0.29 & 0.92 & 0.11 & 0.03 & \\
 & 3 & 0.50 & 3.12 & 11 & 0.77 & TRUE & 0.28 & 0.92 & 0.11 & 0.01 & \\
 \\
 \\[1mm]
\arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{2}{*}{ 1.40} 
 & 1 & 1.00 & 3.20 & 6 & 1.02 & TRUE & 0.28 & 0.56 & 0.14 & 0.01 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{fig9_diagnostic_7.png}}\\
 & 2 & 0.52 & 3.27 & 11 & 1.23 & TRUE & 0.28 & 0.57 & 0.14 & 0.00 & \\
 \\
 \\
 \\[1mm]
 \arrayrulecolor{lightgray}\hline \\[-1mm]
\multirow{2}{*}{ 2.94} 
 & 1 & 0.50 & 1.60 & 6 & 1.19 & TRUE & 0.34 & 0.45 & -0.10 & 0.06 & \multirow{5}{*}{ \includegraphics[scale=0.20, trim = 0mm 10mm 0mm 3mm, clip]{fig9_diagnostic_8.png}}\\
 & 2 & 0.61 & 3.81 & 11 & 2.58 & TRUE & 0.34 & 0.45 & -0.07 & 0.01 & \\
 \\
 \\
 \\[1mm]
\arrayrulecolor{gray}\hline
\end{tabular}
\caption{diagnostic on various data sets with varying smoothness. RMSE* refers to the root mean squared error computed with the exact GP model as the reference.}
  \label{tab1_diagnostic}
\end{table*}

@@

\begin{table}
\centering
\setlength{\tabcolsep}{4pt}
\begin{tabular}{ccccccccc}
\arrayrulecolor{gray}\hline \\[-3mm]
iter. & f & $\ell$ & $c$ & $m$ & $\hat{\ell}$ & $\hat{\ell} + 0.01 \geq \ell$ & RMSE & ELPD \\ 
\arrayrulecolor{lightgray}\hline \\[-2mm]
   & $f_1$ & 0.52 & 1.20 & 7  & 0.53 & FALSE &      &  	   \\
 1 & $f_2$ & 0.50 & -    & 8  & 0.34 & FALSE & 0.32 & 0.05 \\
   & $f_3$ & 0.50 & -    & 8  & 1.62 & TRUE  &      &  	   \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
   & $f_1$ & 0.50 & 1.20 & 8  & 0.54 & TRUE  &      &  	   \\
 2 & $f_2$ & 0.34 & -    & 11 & 0.29 & FALSE & 0.32 & 0.06 \\
   & $f_3$ & 0.29 & -    & 13 & 1.53 & TRUE  &      &  	   \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
   & $f_1$ & 0.28 & 1.20 & 13 & 0.33 & TRUE  &      &  	   \\
 3 & $f_2$ & 0.29 & -    & 13 & 0.24 & FALSE & 0.30 & 0.12 \\
   & $f_3$ & 0.29 & -    & 13 & 1.64 & TRUE  &      &  	   \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
   & $f_1$ & 0.28 & 1.20 & 18 & 0.23 & TRUE  &      &  	   \\
 4 & $f_2$ & 0.24 & -    & 16 & 0.23 & TRUE  & 0.30 & 0.15 \\
   & $f_3$ & 0.29 & -    & 13 & 1.68 & TRUE  &      &  	   \\
\arrayrulecolor{lightgray}\hline \\[-2mm]
   & $f_1$ & 0.16 & 1.20 & 23 & 0.20 & TRUE  &      &  	   \\
 5 & $f_2$ & 0.24 & -    & 16 & 0.23 & TRUE  & 0.29 & 0.18 \\
   & $f_3$ & 0.29 & -    & 13 & 1.63 & TRUE  &      &  	   \\[1mm]
\arrayrulecolor{gray}\hline
\end{tabular}
\caption{Iterative steps in HSGP model fitting of \textit{Birthday data} case study. The diagnosis procedure updating the parameters minimum $\ell$, $c$, $m$, estimated $\hat{\ell}$ and diagnostic $\hat{\ell} + 0.01 \geq \ell$ is applied on the three underlying functions $f_1$, $f_2$ and $f_3$. Notice that $f_2$ and $f_3$ do not have boundary factor $c$ as they use periodic covariance functions. RMSE and ELPD are evaluated for the model underlying function $\mu=f_1 + f_2 + f_3 + f_4$ and sample data. The full diagnosis process is applied until two \textit{trues} are achieved for each function.}
  \label{tab_caseI}
\end{table}
